
# Reproducible Research Peer Assignment - Markdown document. 

## Loading Libraries
```{r Libraries, echo=TRUE}
library(lattice)
library(plyr)
```

## Loading and preprocessing the data
```{r SourceData, echo=TRUE }
act<-read.csv("activity.csv",header=TRUE,sep=",")
summary(act)
dim(act)
names(act)

#changing the class of the variable "steps" to be Numeric:
act$steps <- as.numeric(act$steps) 

#changing the class of the variable "interval" to be Numeric:
act$interval <- as.numeric(act$interval) 

Actuals<- act[complete.cases(act),] # removing NAs.
```

## What is mean total number of steps taken per day?


### Make a histogram of the total number of steps taken each day
```{r Histogram,echo=TRUE}
smsteps<- ddply(Actuals, c("date"), function (x) apply(x[1], 2, sum))

hist(smsteps$steps,main="Histogram of Total Number of Steps per Day",xlab="Total Number of Steps per Day",ylab="Frequency",col='red')

```
### Calculate and report the mean and median total number of steps taken per day 
```{r mean,echo=TRUE}

# mean of the steps
mean(smsteps$steps,na.rm=TRUE) 

# median of the steps
median(smsteps$steps,na.rm=TRUE)
```

## What is the average daily activity pattern?

### Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r avg,echo=TRUE }

# Mean of the steps related to each interval
mnsteps<- ddply(Actuals, c("interval"), function (x) apply(x[1], 2, mean))

xyplot(mnsteps$steps ~ mnsteps$interval,type= "l", ylab="Mean Steps",
       xlab= "Intervals", main="Plot of Intervals vs. Mean Steps",lwd=1.0)
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r maxsteps,echo=TRUE}
mnsteps[mnsteps$steps==max(mnsteps$steps),]

```

## Imputing missing values

### Calculate and report the total number of missing values in the dataset

```{r missing values,echo=TRUE}

# number of missing rows in column 'steps'
sum(is.na(act$steps))

# number of missing rows in column 'interval'
sum(is.na(act$interval))

# number of missing rows in column 'date'
sum(is.na(act$date))

#Replacing the steps of the rows having NAs with the mean steps for each 5-min interval across the all data:

# New dataset but with the missing values filled in

newData<- act
newData[1:288, 1] <- mnsteps$steps 
newData[2017:2304, 1]<- mnsteps$steps
newData[8929:9216, 1]<- mnsteps$steps
newData[9793:10080, 1]<- mnsteps$steps
newData[11233:11520, 1]<- mnsteps$steps
newData[11521:11808, 1]<- mnsteps$steps
newData[12673:12960, 1]<- mnsteps$steps
newData[17281:17568, 1]<- mnsteps$steps

sum(is.na(newData)) # There is no NA.

newSum<- ddply(newData, c("date"), function (x) apply(x[1], 2, sum))

# Plot histogram with new set of activity data 

hist(newSum$steps,main="Histogram of Steps per Day",xlab="Total No of Steps per Day",ylab="Frequency",col='blue')

# mean of the steps
newMean<-mean(newSum$steps) 
newMean #new mean with missing values fill in
newMedian<-median(newSum$steps)
newMedian # median of the steps with missing values fill in

# there is no difference in new mean(missing values filled in) and old mean (with missing values)
# There is slight change in new median(missing values filled in)
```

##Are there differences in activity patterns between weekdays and weekends?


```{r Weekdays & Weekends, echo=TRUE}
# Add a new column name of the day(weekday or weekend)

newData$WD<-weekdays(as.Date(newData$date))

newData$WD[newData$WD=="Monday"]="Weekday"
newData$WD[newData$WD=="Tuesday"]="Weekday"
newData$WD[newData$WD=="Wednesday"]="Weekday"
newData$WD[newData$WD=="Thursday"]="Weekday"
newData$WD[newData$WD=="Fridayday"]="Weekday"
newData$WD[newData$WD=="Friday"]="Weekday"
newData$WD[newData$WD=="Saturday"]="Weekend"
newData$WD[newData$WD=="Sunday"]="Weekend"

table(newData$WD)

new_mnsteps<- ddply(newData, c("interval","WD"), function (x) apply(x[1], 2, mean))

xyplot(new_mnsteps$steps ~ new_mnsteps$interval|new_mnsteps$WD,type="l",ylab="Mean", xlab="Interval",layout=c(1,2))

```
